---
title: "Introduction to celliverse"
subtitle: "EWCSR-based clustering, marker discovery, and cell type annotation for single-cell data"
author: '`r paste("Adrian Salavaty", strftime(Sys.time(), format="%d-%b-%Y"))`'
output:
  rmarkdown::html_vignette:
    css: celliverseVignette.css
    toc: true
    toc_depth: 2
    df_print: paged
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Introduction to celliverse}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
body {
  text-align: justify;
}
</style>

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5.5,
  out.width = "90%"
)
options(width = 95)
set.seed(2025)
```

```{r}
spaces <- function(n) {
  paste(rep("&nbsp;", n), collapse = "")
}
```

# Overview

**celliverse** is an R package for single-cell RNA-seq analysis that introduces a family of methods based on **Expression-Weighted Centered Scaled Ranks (EWCSR)**.

The main goals of the package are:

- High-quality unsupervised clustering (even on large datasets)
- Discovery of specific positive and negative markers
- Marker-based cluster evaluation and refinement
- Cell type annotation using curated marker databases

## Core workflows

The core functions can be grouped into four main workflows:

### 1. Clustering and initial marker discovery

- `clustoCell()`

### 2. Marker ranking and cluster refinement

- `markoCell()`
- `markoClust()`

### 3. Cell type annotation

- `typoClust()`
- `typoClustVis()`

### 4. Helper and visualization functions

- `addClustoData()`
- `addTypoData()`
- `getDatasetMarkers()`
- `markerPurity()`
- `signatureDotHeatmap()`
- `clustoCell_TransferLabel()`

## Example workflow

We will demonstrate the full workflow using the well-known **pbmc3k** dataset (3,000 PBMCs).

## Loading packages and data

```{r}
library(celliverse)
library(Seurat)
library(patchwork)   # for combining plots
library(ggplot2)

# Load the built-in pbmc3k sparse count matrix
data("pbmc3k", package = "celliverse")
```

Create a minimal Seurat object (we keep it very light for the vignette)

```{r}
so <- CreateSeuratObject(
  counts = pbmc3k,
  project = "pbmc3k",
  min.cells = 3,
  min.features = 200
)

so[["percent.mt"]] <- PercentageFeatureSet(so, pattern = "^MT-")

so <- subset(so, subset = nFeature_RNA > 200 & nFeature_RNA < 3500 & percent.mt < 12)

so <- NormalizeData(so)
so <- FindVariableFeatures(so, nfeatures = 2000)
so <- ScaleData(so)
so <- RunPCA(so, npcs = 30)
```

We will not run the default Seurat clustering here; instead, we demonstrate **celliverse** clustering from the beginning.

## Clustering with EWCSR — `clustoCell()`

`clustoCell()` is the main entry point of the package. It performs the following steps:

- (Optional) highly variable gene selection  
- EWCSR transformation  
- Leiden clustering  
- Marker discovery  
- (Optional) sub-clustering 

```{r}
cc_pbmc <- clustoCell(
  so              = so,
  assay           = "RNA",
  layer           = "data",           # we use normalized data
  norm_assay      = "RNA",
  norm_layer      = "data",
  
  subset_to_HVG   = TRUE,
  hvg_selection.method = "vst",
  hvg_var_thresh  = 1.0,
  
  log1p           = TRUE,
  
  identify_subclusters = TRUE,
  leiden_resolution    = 0.9,
  subcluster_resolution_weight = 0.75,
  
  sketch          = FALSE,            # set TRUE for >50–100k cells
  # sketch_ncells = 8000,
  
  seed            = 2025,
  verbose         = TRUE
)
```

Check what we received

```{r}
class(cc_pbmc)
names(cc_pbmc)

# Major clusters
table(cc_pbmc$major_clusters)

# Sub-clusters (if any were found)
table(cc_pbmc$sub_clusters, useNA = "ifany")
```

Add cluster labels back to Seurat object

```{r}
so <- addClustoData(
  obj         = so,
  clustoCell  = cc_pbmc,
  add_major_clusters = TRUE,
  add_sub_clusters   = TRUE,
  major_cluster_name = "ClustoCell",
  sub_cluster_name   = "ClustoCell_sub"
)

DimPlot(so, group.by = "ClustoCell", label = TRUE, repel = TRUE) +
  NoLegend() +
  ggtitle("celliverse major clusters")

DimPlot(so, group.by = "ClustoCell_sub", label = TRUE, repel = TRUE) +
  NoLegend() +
  ggtitle("celliverse sub-clusters (if any)")
```

Extracting and inspecting markers

```{r}

markers <- getDatasetMarkers(
  obj = cc_pbmc,
  clusters       = TRUE,
  sub_clusters   = TRUE,
  positive_markers = TRUE,
  negative_markers = TRUE,
  medium_markers   = FALSE,
  thresh_mode    = "n",
  pos_thresh     = 30,
  neg_thresh     = 20
)

class(markers)
names(markers)

```

# Cell type annotation — typoClust()

We now try to annotate the clusters using curated marker databases.

```{r}
tc_pbmc <- typoClust(
  objects     = list(cc_pbmc),
  tissue      = c("Blood"),
  condition   = c("Healthy"),
  mode        = "markerDB",
  species     = "human",
  thresh_mode = "n",
  thresh      = 25,
  use_pos_markers = TRUE,
  use_neg_markers = TRUE,
  verbose     = TRUE
)

class(tc_pbmc)
names(tc_pbmc)
```

Visualize annotation results

```{r}
p_anno <- typoClustVis(
  typoClust   = tc_pbmc,
  rank_thresh = 1,
  refine      = TRUE,
  refine_thresh = 1,
  order_by    = "Combined Score",
  flip        = FALSE,
  title       = "celliverse cell type annotation — pbmc3k"
)

print(p_anno)
```

Add best cell type labels to Seurat object

```{r}
so <- addTypoData(
  obj       = so,
  typoClust = tc_pbmc,
  clusters  = "ClustoCell",
  rank_thresh = 1,
  refine      = TRUE,
  refine_thresh = 2,
  outNames    = "celliverse_celltype"
)

```

# Summary — recommended workflow

1. Normalize & HVG selection (Seurat / sctransform / ...)

   ↓

2. clustoCell(..., subset_to_HVG = TRUE, identify_subclusters = TRUE)
   → major + sub-clusters + markers

   ↓

3. addClustoData()   → put clusters into Seurat/SCE object

   ↓

4. typoClust()       → cell type annotation

   ↓

5. addTypoData()     → add cell type labels

   ↓

6. visualization:
   • DimPlot( group.by = "celliverse_celltype" )
   • typoClustVis()
   • signatureDotHeatmap()